---
title: "EDA"
author: "Zach Ginder"
format: html
editor: visual
---

## Introduction

The data explored in this analysis comes from the the CDC's BRFSS2015 survey. The target variable for analysis is the Diabetes_binary variable which is in 2 classes: no diabetes or prediabetes/diabetes. There are 21 predictor variables in the dataset to potentially include in the subsequent models.

+-----------------------+------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------+-----------------------------------------------------------------------------------------+
| Variable Name         | Description                                                                                                                                                                                                                  | Classes or Range                                                                        |
+=======================+==============================================================================================================================================================================================================================+=========================================================================================+
| HighBP                | Factor variable for high blood pressure                                                                                                                                                                                      | "No High BP" or "High BP"                                                               |
+-----------------------+------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------+-----------------------------------------------------------------------------------------+
| HighChol              | Factor variable for high cholesterol                                                                                                                                                                                         | "No High Cholesterol" or "High Cholesterol"                                             |
+-----------------------+------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------+-----------------------------------------------------------------------------------------+
| CholCheck             | Factor variable for if the respondent received a cholesterol check in the last 5 years                                                                                                                                       | "No Cholesterol Check In 5 Years" or "Yes Cholesterol Check In 5 Years"                 |
+-----------------------+------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------+-----------------------------------------------------------------------------------------+
| BMI                   | Continuous variable for the body mass index of the respondent                                                                                                                                                                | Values greater than 0                                                                   |
+-----------------------+------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------+-----------------------------------------------------------------------------------------+
| Smoker                | Factor variable where a smoker means the respondent has smoked at least 100 cigarettes (5 packs) in their entire lifetime.                                                                                                   | "Smoker" or "Not A Smoker"                                                              |
+-----------------------+------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------+-----------------------------------------------------------------------------------------+
| Stroke                | Factor variable for whether the respondent has been diagnosed with a stroke or not.                                                                                                                                          | "Diagnosed With Stroke" or "No Stroke"                                                  |
+-----------------------+------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------+-----------------------------------------------------------------------------------------+
| HeartDiseaseorAttack  | Factor variable for whether the respondent has been diagnosed with heart disease and/or myocardial infarction or not.                                                                                                        | "Coronary Heart Disease or Myocardial Infarction" or "No Heart Disease Evident"         |
+-----------------------+------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------+-----------------------------------------------------------------------------------------+
| PhysActivity          | Factor variable for whether the respondent was physically active in the past 30 days or not (not including job).                                                                                                             | "Physical Activity in Past 30 Days" or "No Physical Activity in Past 30 Days"           |
+-----------------------+------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------+-----------------------------------------------------------------------------------------+
| Fruits                | Factor variable on whether the respondent consumes fruit regularly, defined as consuming fruit 1 or more times per day.                                                                                                      | "Consume Fruit 1 Or More Times Per Day" or "Does Not Consume Fruits Regularly"          |
+-----------------------+------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------+-----------------------------------------------------------------------------------------+
| Veggies               | Factor variable on whether the respondent consumes vegetables regularly, defined as consuming vegetables 1 or more times per day.                                                                                            | "Consume Vegetables 1 Or More Times Per Day" or "Does Not Consume Vegetables Regularly" |
+-----------------------+------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------+-----------------------------------------------------------------------------------------+
| HvyAlcoholConsump     | Factor variable on whether the respondent is a heavy alcohol consumer or not. This is defined as heavy if the respondent is an adult men and drinks 14 or more drinks per week or 7 or more drinks per week for adult women. | "Heavy Alcohol Consumer" or "Not Heavy Alcohol Consumer"                                |
+-----------------------+------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------+-----------------------------------------------------------------------------------------+
| AnyHealthcare         | Factor variable for whether the respondent has any kind of health care coverage, including health insurance, prepaid plans such as HMO, etc.                                                                                 | "Has Health Insurance" or "No Health Insurance"                                         |
+-----------------------+------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------+-----------------------------------------------------------------------------------------+
| NoDocbcCost           | Factor variable for whether the respondent in the past 12 months could not see a doctor due to cost but needed to.                                                                                                           | "Missed Doctors Visit Due to Cost" or "No Cost Barrier to Healthcare"                   |
+-----------------------+------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------+-----------------------------------------------------------------------------------------+
| GenHlth               | Factor variable for the respondents general health.                                                                                                                                                                          | "Excellent"                                                                             |
|                       |                                                                                                                                                                                                                              |                                                                                         |
|                       |                                                                                                                                                                                                                              | "Very Good"                                                                             |
|                       |                                                                                                                                                                                                                              |                                                                                         |
|                       |                                                                                                                                                                                                                              | "Good"                                                                                  |
|                       |                                                                                                                                                                                                                              |                                                                                         |
|                       |                                                                                                                                                                                                                              | "Fair"                                                                                  |
|                       |                                                                                                                                                                                                                              |                                                                                         |
|                       |                                                                                                                                                                                                                              | "Poor"                                                                                  |
+-----------------------+------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------+-----------------------------------------------------------------------------------------+
| MentHlth              | Count variable for the number of days of poor mental health in the last 30 days.                                                                                                                                             | Scale from 1 to 30                                                                      |
+-----------------------+------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------+-----------------------------------------------------------------------------------------+
| PhysHlth              | Count variable for the number of days of physical illness or injury in the last 30 days.                                                                                                                                     | Scale from 1 to 30                                                                      |
+-----------------------+------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------+-----------------------------------------------------------------------------------------+
| DiffWalk              | Factor variable for whether the respondent had difficulty with walking or climbing stairs.                                                                                                                                   | "Difficulty With Walking or Stairs" or "No Difficulty With Walking or Stairs"           |
+-----------------------+------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------+-----------------------------------------------------------------------------------------+
| Sex                   | Factor variable for the respondents sex.                                                                                                                                                                                     | "Female" or "Male"                                                                      |
+-----------------------+------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------+-----------------------------------------------------------------------------------------+
| Age                   | Factor variable for the respondents age.                                                                                                                                                                                     | "18-24"                                                                                 |
|                       |                                                                                                                                                                                                                              |                                                                                         |
|                       |                                                                                                                                                                                                                              | "25-29"                                                                                 |
|                       |                                                                                                                                                                                                                              |                                                                                         |
|                       |                                                                                                                                                                                                                              | "30-34"                                                                                 |
|                       |                                                                                                                                                                                                                              |                                                                                         |
|                       |                                                                                                                                                                                                                              | "35-39"                                                                                 |
|                       |                                                                                                                                                                                                                              |                                                                                         |
|                       |                                                                                                                                                                                                                              | "40-44"                                                                                 |
|                       |                                                                                                                                                                                                                              |                                                                                         |
|                       |                                                                                                                                                                                                                              | "45-49"                                                                                 |
|                       |                                                                                                                                                                                                                              |                                                                                         |
|                       |                                                                                                                                                                                                                              | "50-54"                                                                                 |
|                       |                                                                                                                                                                                                                              |                                                                                         |
|                       |                                                                                                                                                                                                                              | "55-59"                                                                                 |
|                       |                                                                                                                                                                                                                              |                                                                                         |
|                       |                                                                                                                                                                                                                              | "60-64"                                                                                 |
|                       |                                                                                                                                                                                                                              |                                                                                         |
|                       |                                                                                                                                                                                                                              | "65-69"                                                                                 |
|                       |                                                                                                                                                                                                                              |                                                                                         |
|                       |                                                                                                                                                                                                                              | "70-74"                                                                                 |
|                       |                                                                                                                                                                                                                              |                                                                                         |
|                       |                                                                                                                                                                                                                              | "75-79"                                                                                 |
|                       |                                                                                                                                                                                                                              |                                                                                         |
|                       |                                                                                                                                                                                                                              | "80 or older"                                                                           |
+-----------------------+------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------+-----------------------------------------------------------------------------------------+
| Education             | Factor variable for the respondents highest completed education.                                                                                                                                                             | "Never Attended School Or Only Kindergarten"                                            |
|                       |                                                                                                                                                                                                                              |                                                                                         |
|                       |                                                                                                                                                                                                                              | "Grades 1-8"                                                                            |
|                       |                                                                                                                                                                                                                              |                                                                                         |
|                       |                                                                                                                                                                                                                              | "Grades 9-11"                                                                           |
|                       |                                                                                                                                                                                                                              |                                                                                         |
|                       |                                                                                                                                                                                                                              | "Grades 12 or GED"                                                                      |
|                       |                                                                                                                                                                                                                              |                                                                                         |
|                       |                                                                                                                                                                                                                              | "Some College or Technical School"                                                      |
|                       |                                                                                                                                                                                                                              |                                                                                         |
|                       |                                                                                                                                                                                                                              | "Graduated College or Technical School"                                                 |
+-----------------------+------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------+-----------------------------------------------------------------------------------------+
| Income                | Factor variable for the respondents income.                                                                                                                                                                                  | "Less than \$10,000"                                                                    |
|                       |                                                                                                                                                                                                                              |                                                                                         |
|                       |                                                                                                                                                                                                                              | "\$10,000 to less than \$15,000"                                                        |
|                       |                                                                                                                                                                                                                              |                                                                                         |
|                       |                                                                                                                                                                                                                              | "\$15,000 to less than \$20,000"                                                        |
|                       |                                                                                                                                                                                                                              |                                                                                         |
|                       |                                                                                                                                                                                                                              | "\$20,000 to less than \$25,000"                                                        |
|                       |                                                                                                                                                                                                                              |                                                                                         |
|                       |                                                                                                                                                                                                                              | "\$25,000 to less than \$35,000"                                                        |
|                       |                                                                                                                                                                                                                              |                                                                                         |
|                       |                                                                                                                                                                                                                              | "\$35,000 to less than \$50,000"                                                        |
|                       |                                                                                                                                                                                                                              |                                                                                         |
|                       |                                                                                                                                                                                                                              | "\$50,000 to less than \$75,000"                                                        |
|                       |                                                                                                                                                                                                                              |                                                                                         |
|                       |                                                                                                                                                                                                                              | "\$75,000 or more"                                                                      |
+-----------------------+------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------+-----------------------------------------------------------------------------------------+

The dataset is unbalanced meaning there are more cases where the survey respondent had no diabetes than in the case where the survey respondent either had prediabetes or diabetes. This is important when it comes to modeling as we likely will want to ensure that our model does better than the majority class rate, or rate of the largest class of the dataset, in this case being the no diabetes class. If a model only predicted the majority class it would still achieve this majority class accuracy, but it would not be useful. Ensuring the model is useful in predicting the Diabetes_binary variable based on a set of predictors is the goal and ensuring the accuracy of the model is better than the majority class rate will be one important evaluation of model importance. This exploratory data analysis serves to try to determine a few useful predictors to use when

## Data

```{r}
#Library these packages
library(tidyverse)

#Read in csv file
Diabetes<-read.csv("diabetes_binary_health_indicators_BRFSS2015.csv")

#Determine structure of data
str(Diabetes)
```

The data for this analysis can be found at the following website (<https://www.kaggle.com/datasets/alexteboul/diabetes-health-indicators-dataset/>). As noticed above, all the variables are treated as numeric; however, a majority of the variables are actually factors. We can transform those variables into factors with meaningful level names.

```{r}
#Transform data from numeric to factor
Diabetes<-Diabetes|>
  mutate(Diabetes_binary=factor(Diabetes_binary,levels=c("0","1"),
                                labels=c("No Diabetes","Prediabetes/Diabetes")),
         HighBP=factor(HighBP,levels=c("0","1"),
                          labels=c("No High BP","High BP")),
         HighChol=factor(HighChol,levels=c("0","1"),
                            labels=c("No High Cholesterol","High Cholesterol")),
         CholCheck=factor(CholCheck,levels=c("0","1"),
                             labels=c("No Cholesterol Check In 5 Years",
                                      "Yes Cholesterol Check In 5 Years")),
         Smoker=factor(Smoker,levels=c("1","0"),
                             labels=c("Smoker",
                                      "Not A Smoker")),
         Stroke=factor(Stroke,levels=c("1","0"),
                             labels=c("Diagnosed With Stroke","No Stroke")),
         HeartDiseaseorAttack=factor(HeartDiseaseorAttack,levels=c("1","0"),
                             labels=c("Coronary Heart Disease or Myocardial Infarction",
                                      "No Heart Disease Evident")),
         PhysActivity=factor(PhysActivity,levels=c("1","0"),
                             labels=c("Physical Activity in Past 30 Days",
                                      "No Physical Activity in Past 30 Days")),
         Fruits=factor(Fruits,levels=c("1","0"),
                             labels=c("Consume Fruit 1 Or More Times Per Day",
                                      "Does Not Consume Fruits Regularly")),
         Veggies=factor(Veggies,levels=c("1","0"),
                             labels=c("Consume Vegetables 1 Or More Times Per Day",
                                      "Does Not Consume Vegetables Regularly")),
         HvyAlcoholConsump=factor(HvyAlcoholConsump,levels=c("1","0"),
                             labels=c("Heavy Alcohol Consumer",
                                      "Not Heavy Alcohol Consumer")),
         AnyHealthcare=factor(AnyHealthcare,levels=c("1","0"),
                             labels=c("Has Health Insurance",
                                      "No Health Insurance")),
         NoDocbcCost=factor(NoDocbcCost,levels=c("1","0"),
                             labels=c("Missed Doctors Visit Due to Cost",
                                      "No Cost Barrier to Healthcare")),
         GenHlth=factor(GenHlth,levels=c("1","2","3","4","5"),
                             labels=c("Excellent","Very Good","Good","Fair","Poor")),
         DiffWalk=factor(DiffWalk,levels=c("1","0"),
                             labels=c("Difficulty With Walking or Stairs",
                                      "No Difficulty With Walking or Stairs")),
         Sex=factor(Sex,levels=c("0","1"),
                             labels=c("Female","Male")),
         Age=factor(Age,levels=c("1","2","3","4","5","6","7","8","9","10",
                                         "11","12","13"),
                             labels=c("18-24","25-29","30-34","35-39","40-44",
                                      "45-49","50-54","55-59","60-64","65-69",
                                      "70-74","75-79","80 or older")),
         Education=factor(Education,levels=c("1","2","3","4","5","6"),
                             labels=c("Never Attended School Or Only Kindergarten",
                                      "Grades 1-8",
                                      "Grades 9-11",
                                      "Grades 12 or GED",
                                      "Some College or Technical School",
                                      "Graduated College or Technical School")),
         Income=factor(Income,levels=c("1","2","3","4","5","6","7","8"),
                             labels=c("Less than $10,000",
                                      "$10,000 to less than $15,000",
                                      "$15,000 to less than $20,000",
                                      "$20,000 to less than $25,000",
                                      "$25,000 to less than $35,000",
                                      "$35,000 to less than $50,000",
                                      "$50,000 to less than $75,000",
                                      "$75,000 or more")),
         )
```

After transforming the data so that appropriate categorical variables are stored as such, it is important to check for missing values in the data as that can cause issues with modeling.

```{r}
#Check for na's in the data set Diabetes
any(is.na(Diabetes))
```

The result of FALSE shows there are no NA's in the data set. This means this data is ready for the exploratory data analysis stage.

## Summarizations

In this exploratory data analysis stage, we want to evaluate three variables that may be beneficial in predicting the diabetes status (i.e. the Diabetes_binary variable). To do this we can explore the variables in both uni variate and bivariate analyses.

```{r}
#Evaluate the response variable: Diabetes_binary
barplot(table(Diabetes$Diabetes_binary),ylim=c(0,250000))
grid(nx=NA,ny=NULL,lty=3,col="black")

table(Diabetes$Diabetes_binary)
```

```{r}
#Evaluate the HighBP variable
table(Diabetes_binary=Diabetes$Diabetes_binary,HighBP=Diabetes$HighBP)
chisq.test(table(Diabetes_binary=Diabetes$Diabetes_binary,HighBP=Diabetes$HighBP))
```

Using a Chi-square test for independence it appears statistically significant suggesting that the variables of HighBP and Diabetes_binary are not independent, and thus should be considered for use as a predictor. I will do this same procedure for all additional categorical/factor variables next.

```{r}
#| echo: false
#Evaluate the HighChol variable
table(Diabetes_binary=Diabetes$Diabetes_binary,HighChol=Diabetes$HighChol)
chisq.test(table(Diabetes_binary=Diabetes$Diabetes_binary,HighChol=Diabetes$HighChol))

#Evaluate the CholCheck variable
table(Diabetes_binary=Diabetes$Diabetes_binary,CholCheck=Diabetes$CholCheck)
chisq.test(table(Diabetes_binary=Diabetes$Diabetes_binary,CholCheck=Diabetes$CholCheck))

#Evaluate the Smoker variable
table(Diabetes_binary=Diabetes$Diabetes_binary,Smoker=Diabetes$Smoker)
chisq.test(table(Diabetes_binary=Diabetes$Diabetes_binary,Smoker=Diabetes$Smoker))

#Evaluate the Stroke variable
table(Diabetes_binary=Diabetes$Diabetes_binary,Stroke=Diabetes$Stroke)
chisq.test(table(Diabetes_binary=Diabetes$Diabetes_binary,Stroke=Diabetes$Stroke))

#Evaluate the HeartDiseaseorAttack variable
table(Diabetes_binary=Diabetes$Diabetes_binary,
      HeartDiseaseorAttack=Diabetes$HeartDiseaseorAttack)
chisq.test(table(Diabetes_binary=Diabetes$Diabetes_binary,
                 HeartDiseaseorAttack=Diabetes$HeartDiseaseorAttack))

#Evaluate the PhysActivity variable
table(Diabetes_binary=Diabetes$Diabetes_binary,
      PhysActivity=Diabetes$PhysActivity)
chisq.test(table(Diabetes_binary=Diabetes$Diabetes_binary,
                 PhysActivity=Diabetes$PhysActivity))

#Evaluate the Fruits variable
table(Diabetes_binary=Diabetes$Diabetes_binary,
      Fruits=Diabetes$Fruits)
chisq.test(table(Diabetes_binary=Diabetes$Diabetes_binary,
                 Fruits=Diabetes$Fruits))

#Evaluate the Veggies variable
table(Diabetes_binary=Diabetes$Diabetes_binary,
      Veggies=Diabetes$Veggies)
chisq.test(table(Diabetes_binary=Diabetes$Diabetes_binary,
                 Veggies=Diabetes$Veggies))

#Evaluate the HvyAlcoholConsump variable
table(Diabetes_binary=Diabetes$Diabetes_binary,
      HvyAlcoholConsump=Diabetes$HvyAlcoholConsump)
chisq.test(table(Diabetes_binary=Diabetes$Diabetes_binary,
                 HvyAlcoholConsump=Diabetes$HvyAlcoholConsump))

#Evaluate the AnyHealthcare variable
table(Diabetes_binary=Diabetes$Diabetes_binary,
      AnyHealthcare=Diabetes$AnyHealthcare)
chisq.test(table(Diabetes_binary=Diabetes$Diabetes_binary,
                 AnyHealthcare=Diabetes$AnyHealthcare))

#Evaluate the NoDocbcCost variable
table(Diabetes_binary=Diabetes$Diabetes_binary,
      NoDocbcCost=Diabetes$NoDocbcCost)
chisq.test(table(Diabetes_binary=Diabetes$Diabetes_binary,
                 NoDocbcCost=Diabetes$NoDocbcCost))

#Evaluate the GenHlth variable
table(Diabetes_binary=Diabetes$Diabetes_binary,
      GenHlth=Diabetes$GenHlth)
chisq.test(table(Diabetes_binary=Diabetes$Diabetes_binary,
                 GenHlth=Diabetes$GenHlth))

#Evaluate the DiffWalk variable
table(Diabetes_binary=Diabetes$Diabetes_binary,
      DiffWalk=Diabetes$DiffWalk)
chisq.test(table(Diabetes_binary=Diabetes$Diabetes_binary,
                 DiffWalk=Diabetes$DiffWalk))

#Evaluate the Sex variable
table(Diabetes_binary=Diabetes$Diabetes_binary,
      Sex=Diabetes$Sex)
chisq.test(table(Diabetes_binary=Diabetes$Diabetes_binary,
                 Sex=Diabetes$Sex))

#Evaluate the Age variable
table(Diabetes_binary=Diabetes$Diabetes_binary,
      Age=Diabetes$Age)
chisq.test(table(Diabetes_binary=Diabetes$Diabetes_binary,
                 Age=Diabetes$Age))

#Evaluate the Education variable
table(Diabetes_binary=Diabetes$Diabetes_binary,
      Education=Diabetes$Education)
chisq.test(table(Diabetes_binary=Diabetes$Diabetes_binary,
                 Education=Diabetes$Education))

#Evaluate the Income variable
table(Diabetes_binary=Diabetes$Diabetes_binary,
      Income=Diabetes$Income)
chisq.test(table(Diabetes_binary=Diabetes$Diabetes_binary,
                 Income=Diabetes$Income))
```
