---
title: "EDA"
author: "Zach Ginder"
format: html
editor: visual
---

## Introduction

The data explored in this analysis comes from the the CDC's BRFSS2015 survey. The target variable for analysis is the Diabetes_binary variable which is in 2 classes: no diabetes or prediabetes/diabetes. There are 21 predictor variables in the dataset to potentially include in the subsequent models.

+----------------------+------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------+-----------------------------------------------------------------------------------------+
| Variable Name        | Description                                                                                                                                                                                                                  | Classes or Range                                                                        |
+======================+==============================================================================================================================================================================================================================+=========================================================================================+
| HighBP               | Factor variable for high blood pressure                                                                                                                                                                                      | "No High BP" or "High BP"                                                               |
+----------------------+------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------+-----------------------------------------------------------------------------------------+
| HighChol             | Factor variable for high cholesterol                                                                                                                                                                                         | "No High Cholesterol" or "High Cholesterol"                                             |
+----------------------+------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------+-----------------------------------------------------------------------------------------+
| CholCheck            | Factor variable for if the respondent received a cholesterol check in the last 5 years                                                                                                                                       | "No Cholesterol Check In 5 Years" or "Yes Cholesterol Check In 5 Years"                 |
+----------------------+------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------+-----------------------------------------------------------------------------------------+
| BMI                  | Continuous variable for the body mass index of the respondent                                                                                                                                                                | Values greater than 0                                                                   |
+----------------------+------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------+-----------------------------------------------------------------------------------------+
| Smoker               | Factor variable where a smoker means the respondent has smoked at least 100 cigarettes (5 packs) in their entire lifetime.                                                                                                   | "Smoker" or "Not A Smoker"                                                              |
+----------------------+------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------+-----------------------------------------------------------------------------------------+
| Stroke               | Factor variable for whether the respondent has been diagnosed with a stroke or not.                                                                                                                                          | "Diagnosed With Stroke" or "No Stroke"                                                  |
+----------------------+------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------+-----------------------------------------------------------------------------------------+
| HeartDiseaseorAttack | Factor variable for whether the respondent has been diagnosed with heart disease and/or myocardial infarction or not.                                                                                                        | "Coronary Heart Disease or Myocardial Infarction" or "No Heart Disease Evident"         |
+----------------------+------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------+-----------------------------------------------------------------------------------------+
| PhysActivity         | Factor variable for whether the respondent was physically active in the past 30 days or not (not including job).                                                                                                             | "Physical Activity in Past 30 Days" or "No Physical Activity in Past 30 Days"           |
+----------------------+------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------+-----------------------------------------------------------------------------------------+
| Fruits               | Factor variable on whether the respondent consumes fruit regularly, defined as consuming fruit 1 or more times per day.                                                                                                      | "Consume Fruit 1 Or More Times Per Day" or "Does Not Consume Fruits Regularly"          |
+----------------------+------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------+-----------------------------------------------------------------------------------------+
| Veggies              | Factor variable on whether the respondent consumes vegetables regularly, defined as consuming vegetables 1 or more times per day.                                                                                            | "Consume Vegetables 1 Or More Times Per Day" or "Does Not Consume Vegetables Regularly" |
+----------------------+------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------+-----------------------------------------------------------------------------------------+
| HvyAlcoholConsump    | Factor variable on whether the respondent is a heavy alcohol consumer or not. This is defined as heavy if the respondent is an adult men and drinks 14 or more drinks per week or 7 or more drinks per week for adult women. | "Heavy Alcohol Consumer" or "Not Heavy Alcohol Consumer"                                |
+----------------------+------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------+-----------------------------------------------------------------------------------------+
| AnyHealthcare        | Factor variable for whether the respondent has any kind of health care coverage, including health insurance, prepaid plans such as HMO, etc.                                                                                 | "Has Health Insurance" or "No Health Insurance"                                         |
+----------------------+------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------+-----------------------------------------------------------------------------------------+
| NoDocbcCost          | Factor variable for whether the respondent in the past 12 months could not see a doctor due to cost but needed to.                                                                                                           | "Missed Doctors Visit Due to Cost" or "No Cost Barrier to Healthcare"                   |
+----------------------+------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------+-----------------------------------------------------------------------------------------+
| GenHlth              | Factor variable for the respondents general health.                                                                                                                                                                          | "Excellent"                                                                             |
|                      |                                                                                                                                                                                                                              |                                                                                         |
|                      |                                                                                                                                                                                                                              | "Very Good"                                                                             |
|                      |                                                                                                                                                                                                                              |                                                                                         |
|                      |                                                                                                                                                                                                                              | "Good"                                                                                  |
|                      |                                                                                                                                                                                                                              |                                                                                         |
|                      |                                                                                                                                                                                                                              | "Fair"                                                                                  |
|                      |                                                                                                                                                                                                                              |                                                                                         |
|                      |                                                                                                                                                                                                                              | "Poor"                                                                                  |
+----------------------+------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------+-----------------------------------------------------------------------------------------+
| MentHlth             | Count variable for the number of days of poor mental health in the last 30 days.                                                                                                                                             | Scale from 1 to 30                                                                      |
+----------------------+------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------+-----------------------------------------------------------------------------------------+
| PhysHlth             | Count variable for the number of days of physical illness or injury in the last 30 days.                                                                                                                                     | Scale from 1 to 30                                                                      |
+----------------------+------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------+-----------------------------------------------------------------------------------------+
| DiffWalk             | Factor variable for whether the respondent had difficulty with walking or climbing stairs.                                                                                                                                   | "Difficulty With Walking or Stairs" or "No Difficulty With Walking or Stairs"           |
+----------------------+------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------+-----------------------------------------------------------------------------------------+
| Sex                  | Factor variable for the respondents sex.                                                                                                                                                                                     | "Female" or "Male"                                                                      |
+----------------------+------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------+-----------------------------------------------------------------------------------------+
| Age                  | Factor variable for the respondents age.                                                                                                                                                                                     | "18-24"                                                                                 |
|                      |                                                                                                                                                                                                                              |                                                                                         |
|                      |                                                                                                                                                                                                                              | "25-29"                                                                                 |
|                      |                                                                                                                                                                                                                              |                                                                                         |
|                      |                                                                                                                                                                                                                              | "30-34"                                                                                 |
|                      |                                                                                                                                                                                                                              |                                                                                         |
|                      |                                                                                                                                                                                                                              | "35-39"                                                                                 |
|                      |                                                                                                                                                                                                                              |                                                                                         |
|                      |                                                                                                                                                                                                                              | "40-44"                                                                                 |
|                      |                                                                                                                                                                                                                              |                                                                                         |
|                      |                                                                                                                                                                                                                              | "45-49"                                                                                 |
|                      |                                                                                                                                                                                                                              |                                                                                         |
|                      |                                                                                                                                                                                                                              | "50-54"                                                                                 |
|                      |                                                                                                                                                                                                                              |                                                                                         |
|                      |                                                                                                                                                                                                                              | "55-59"                                                                                 |
|                      |                                                                                                                                                                                                                              |                                                                                         |
|                      |                                                                                                                                                                                                                              | "60-64"                                                                                 |
|                      |                                                                                                                                                                                                                              |                                                                                         |
|                      |                                                                                                                                                                                                                              | "65-69"                                                                                 |
|                      |                                                                                                                                                                                                                              |                                                                                         |
|                      |                                                                                                                                                                                                                              | "70-74"                                                                                 |
|                      |                                                                                                                                                                                                                              |                                                                                         |
|                      |                                                                                                                                                                                                                              | "75-79"                                                                                 |
|                      |                                                                                                                                                                                                                              |                                                                                         |
|                      |                                                                                                                                                                                                                              | "80 or older"                                                                           |
+----------------------+------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------+-----------------------------------------------------------------------------------------+
| Education            | Factor variable for the respondents highest completed education.                                                                                                                                                             | "Never Attended School Or Only Kindergarten"                                            |
|                      |                                                                                                                                                                                                                              |                                                                                         |
|                      |                                                                                                                                                                                                                              | "Grades 1-8"                                                                            |
|                      |                                                                                                                                                                                                                              |                                                                                         |
|                      |                                                                                                                                                                                                                              | "Grades 9-11"                                                                           |
|                      |                                                                                                                                                                                                                              |                                                                                         |
|                      |                                                                                                                                                                                                                              | "Grades 12 or GED"                                                                      |
|                      |                                                                                                                                                                                                                              |                                                                                         |
|                      |                                                                                                                                                                                                                              | "Some College or Technical School"                                                      |
|                      |                                                                                                                                                                                                                              |                                                                                         |
|                      |                                                                                                                                                                                                                              | "Graduated College or Technical School"                                                 |
+----------------------+------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------+-----------------------------------------------------------------------------------------+
| Income               | Factor variable for the respondents income.                                                                                                                                                                                  | "Less than \$10,000"                                                                    |
|                      |                                                                                                                                                                                                                              |                                                                                         |
|                      |                                                                                                                                                                                                                              | "\$10,000 to less than \$15,000"                                                        |
|                      |                                                                                                                                                                                                                              |                                                                                         |
|                      |                                                                                                                                                                                                                              | "\$15,000 to less than \$20,000"                                                        |
|                      |                                                                                                                                                                                                                              |                                                                                         |
|                      |                                                                                                                                                                                                                              | "\$20,000 to less than \$25,000"                                                        |
|                      |                                                                                                                                                                                                                              |                                                                                         |
|                      |                                                                                                                                                                                                                              | "\$25,000 to less than \$35,000"                                                        |
|                      |                                                                                                                                                                                                                              |                                                                                         |
|                      |                                                                                                                                                                                                                              | "\$35,000 to less than \$50,000"                                                        |
|                      |                                                                                                                                                                                                                              |                                                                                         |
|                      |                                                                                                                                                                                                                              | "\$50,000 to less than \$75,000"                                                        |
|                      |                                                                                                                                                                                                                              |                                                                                         |
|                      |                                                                                                                                                                                                                              | "\$75,000 or more"                                                                      |
+----------------------+------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------+-----------------------------------------------------------------------------------------+

The dataset is unbalanced meaning there are more cases where the survey respondent had no diabetes than in the case where the survey respondent either had prediabetes or diabetes. This is important when it comes to modeling as we likely will want to ensure that our model does better than the majority class rate, or rate of the largest class of the dataset, in this case being the no diabetes class. If a model only predicted the majority class it would still achieve this majority class accuracy, but it would not be useful. Ensuring the model is useful in predicting the Diabetes_binary variable based on a set of predictors is the goal and ensuring the accuracy of the model is better than the majority class rate will be one important evaluation of model importance. This exploratory data analysis serves to try to determine a few useful predictors to use when

## Data

```{r}
#Library these packages
library(tidyverse)

#Read in csv file
Diabetes<-read.csv("diabetes_binary_health_indicators_BRFSS2015.csv")

#Determine structure of data
str(Diabetes)
```

The data for this analysis can be found at the following website (<https://www.kaggle.com/datasets/alexteboul/diabetes-health-indicators-dataset/>). As noticed above, all the variables are treated as numeric; however, a majority of the variables are actually factors. We can transform those variables into factors with meaningful level names.

```{r}
#Transform data from numeric to factor
Diabetes<-Diabetes|>
  mutate(Diabetes_binary=factor(Diabetes_binary,levels=c("0","1"),
                                labels=c("No_Diabetes","Prediabetes_or_Diabetes")),
         HighBP=factor(HighBP,levels=c("0","1"),
                          labels=c("No_High_BP","High_BP")),
         HighChol=factor(HighChol,levels=c("0","1"),
                            labels=c("No_High_Cholesterol","High_Cholesterol")),
         CholCheck=factor(CholCheck,levels=c("0","1"),
                             labels=c("No_Cholesterol_Check_In_5_Years",
                                      "Yes_Cholesterol_Check_In_5_Years")),
         Smoker=factor(Smoker,levels=c("1","0"),
                             labels=c("Smoker",
                                      "Not_A_Smoker")),
         Stroke=factor(Stroke,levels=c("1","0"),
                             labels=c("Diagnosed_With_Stroke","No_Stroke")),
         HeartDiseaseorAttack=factor(HeartDiseaseorAttack,levels=c("1","0"),
                             labels=c("Coronary_Heart_Disease_or_Myocardial_Infarction",
                                      "No_Heart_Disease_Evident")),
         PhysActivity=factor(PhysActivity,levels=c("1","0"),
                             labels=c("Physical_Activity_in_Past_30_Days",
                                      "No_Physical_Activity_in_Past_30_Days")),
         Fruits=factor(Fruits,levels=c("1","0"),
                             labels=c("Consume_Fruit_1_Or_More_Times_Per_Day",
                                      "Does_Not_Consume_Fruits_Regularly")),
         Veggies=factor(Veggies,levels=c("1","0"),
                             labels=c("Consume_Vegetables_1_Or_More_Times_Per_Day",
                                      "Does_Not_Consume_Vegetables_Regularly")),
         HvyAlcoholConsump=factor(HvyAlcoholConsump,levels=c("1","0"),
                             labels=c("Heavy_Alcohol_Consumer",
                                      "Not_Heavy_Alcohol_Consumer")),
         AnyHealthcare=factor(AnyHealthcare,levels=c("1","0"),
                             labels=c("Has_Health_Insurance",
                                      "No_Health_Insurance")),
         NoDocbcCost=factor(NoDocbcCost,levels=c("1","0"),
                             labels=c("Missed_Doctors_Visit_Due_to_Cost",
                                      "No_Cost_Barrier_to_Healthcare")),
         GenHlth=factor(GenHlth,levels=c("1","2","3","4","5"),
                             labels=c("Excellent","Very_Good","Good","Fair","Poor")),
         DiffWalk=factor(DiffWalk,levels=c("1","0"),
                             labels=c("Difficulty_With_Walking_or_Stairs",
                                      "No_Difficulty_With_Walking_or_Stairs")),
         Sex=factor(Sex,levels=c("0","1"),
                             labels=c("Female","Male")),
         Age=factor(Age,levels=c("1","2","3","4","5","6","7","8","9","10",
                                         "11","12","13"),
                             labels=c("18-24","25-29","30-34","35-39","40-44",
                                      "45-49","50-54","55-59","60-64","65-69",
                                      "70-74","75-79","80_or_older")),
         Education=factor(Education,levels=c("1","2","3","4","5","6"),
                             labels=c("Never_Attended_School_Or_Only_Kindergarten",
                                      "Grades_1-8",
                                      "Grades_ 9-11",
                                      "Grades_12_or_GED",
                                      "Some_College_or_Technical_School",
                                      "Graduated_College_or_Technical_School")),
         Income=factor(Income,levels=c("1","2","3","4","5","6","7","8"),
                             labels=c("Less_than_$10,000",
                                      "$10,000_to_less_than_$15,000",
                                      "$15,000_to_less_than_$20,000",
                                      "$20,000_to_less_than_$25,000",
                                      "$25,000_to_less_than_$35,000",
                                      "$35,000_to_less_than_$50,000",
                                      "$50,000_to_less_than_$75,000",
                                      "$75,000_or_more")),
         )
```

After transforming the data so that appropriate categorical variables are stored as such, it is important to check for missing values in the data as that can cause issues with modeling.

```{r}
#Check for na's in the data set Diabetes
any(is.na(Diabetes))
```

The result of FALSE shows there are no NA's in the data set. This means this data is ready for the exploratory data analysis stage.

## Summarizations

In this exploratory data analysis stage, we want to evaluate three variables that may be beneficial in predicting the diabetes status (i.e. the Diabetes_binary variable). To do this we can explore the variables in both uni variate and bivariate analyses.

#### Diabetes_binary

```{r}
#| echo: false
#Evaluate the response variable: Diabetes_binary
barplot(table(Diabetes$Diabetes_binary),ylim=c(0,250000))
grid(nx=NA,ny=NULL,lty=3,col="black")

table(Diabetes$Diabetes_binary)
```

This bar plot shows the frequency of the respondents that had no diabetes diagnosis compared to those that had prediabetes or diabetes diagnoses. This confirms the original detail of the data that this is an unbalanced design. Specifically, we see that $\frac{218334}{253680}\approx$ 86.07%. This is the majority class rate that the accuracy of our models would need to do better than in order to be beneficial.

Three explanatory variables of interest to analyze are BMI, HvyAlcoholConsump, and Income. Research suggests that increased BMI is related to an increase in risk of diabetes (<https://pmc.ncbi.nlm.nih.gov/articles/PMC4457375/>). Furthermore, it is possible that income relates to diabetes risk as healthy food is known to be expensive when compared to cheaper food. Alcohol consumption would also be a consideration as alcohol is high in sugar. Univariate analysis can be performed on these three variables to analyze their distributions. Bivariate analysis between each explanatory variable with the response variable is also useful to determine initial association based on statistical significance.

#### BMI

```{r}
#| echo: false
#Evaluate the explanatory variable: BMI
hist(Diabetes$BMI,xlim=c(0,100),ylim=c(0,100000),main="Histogram of BMI",
     xlab="BMI")
grid(nx=NA,ny=NULL,lty=3,col="black")

#Calculate the mean and range
mean_bmi<-mean(Diabetes$BMI)
range_bmi<-range(Diabetes$BMI)
cat("The median of the BMI of the respondents is", mean_bmi,"\n")
cat("The range of the BMI of the respondents is", range_bmi,"\n")
```

The univariate analysis of the BMI of the survey respondents shows that the distribution of BMIs is right skewed with a median of 28.38236. According to the National Heart, Lung, and Blood Institute, a BMI over 24.9 is considered overweight (<https://www.nhlbi.nih.gov/health/educational/lose_wt/risk.htm>). This analysis shows that a majority of the respondents were overweight based on BMI and earlier research was shown that high BMI is associated with increased risk in diabetes.

```{r}
#| echo: false
boxplot(Diabetes$BMI~Diabetes$Diabetes_binary,
        xlab="Diabetes_binary",
        ylab="BMI",
        outline=FALSE)
```

The bivariate analysis of Diabetes_binary with BMI (with outliers removed) shows that the median BMI for respondents without diabetes is lower than median BMI for respondents with prediabetes/diabetes. This validates the assumption earlier based on the research that BMI may be a useful predictor of diabetes status.

#### Heavy Alcohol Consumption (HvyAlcoholConsump)

```{r}
#| echo: false
#Evaluate the explanatory variable: HvyAlcoholConsump
barplot(table(Diabetes$HvyAlcoholConsump),ylim=c(0,250000))
grid(nx=NA,ny=NULL,lty=3,col="black")

table(Diabetes$HvyAlcoholConsump)
```

The univariate analysis of the HvyAlcoholConsump variable shows a bar plot similar to that of the Diabetes_binary response variable where there is a high frequency in one class and a low frequency in another. As discussed before, alcohol consumption might be associated with diabetes as alcohol contains high levels of sugar. Consider the proportion of respondents with prediabetes/diabetes and those that are heavy alcohol consumers is similar this variable should be considered as a potential explanatory variable.

```{r}
#| echo: false
#Evaluate the HvyAlcoholConsump variable (bivariate analysis)
table(Diabetes_binary=Diabetes$Diabetes_binary,
      HvyAlcoholConsump=Diabetes$HvyAlcoholConsump)
chisq.test(table(Diabetes_binary=Diabetes$Diabetes_binary,
                 HvyAlcoholConsump=Diabetes$HvyAlcoholConsump))
```

The bivariate analysis of Income and Diabetes_binary using a Chi-square test for independence shows statistically significance suggesting that the variables of Income and Diabetes_binary are not independent, and thus should be considered for use as a predictor. We performed this test on the contingency table shown above.

#### Income

```{r}
#| echo: false
#Evaluate the explanatory variable: Income
barplot(table(Diabetes$Income),ylim=c(0,100000))
grid(nx=NA,ny=NULL,lty=3,col="black")

table(Diabetes$Income)
```

The univariate analysis of the Income variable shows that the distribution is left skewed with the income of the respondents being relatively high in terms of the provided classes. As discussed prior, higher incomes are likely associated with less risk of diabetes since higher incomes provide access to healthy foods which are generally lower in sugar. Therefore, this variable may be a useful predictor as similarly to the other variables, the majority class of the variable is associated with the majority class of the diabetes variable which was no diabetes. It can be further analyzed with bivariate analysis whether income is statistically associated with diabetes status.

```{r}
#| echo: false
#Evaluate the Income variable (bivariate analysis)
table(Diabetes_binary=Diabetes$Diabetes_binary,
      Income=Diabetes$Income)
chisq.test(table(Diabetes_binary=Diabetes$Diabetes_binary,
                 Income=Diabetes$Income))
```

Using a Chi-square test for independence it appears statistically significant suggesting that the variables of Income and Diabetes_binary are not independent, and thus should be considered for use as a predictor. We performed this test on the contingency table shown above.

Link to Modeling Page: <https://zmginder.github.io/Project-3/Modeling.html>
